---
title: "Radiometric calibration with reference surfaces using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radiometric calibration with reference surfaces using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

# Introduction

In addition to correcting drift in thermal infrared (TIR) images, the `thermal`
package can also perform radiometric calibration of images or orthomosaics by
combining images with information on the temperature and position of reference
surfaces. This vignette will show a step-by-step procedure to perform such
radiometric calibration, i.e., translating digital number (DN) values returned
by a camera into actual surface temperature readings.

# Dataset

The dataset used here is based on the same flight as the vignette on thermal
drift correction, which we advised to read first. In brief, this dataset
comprises 30 TIR images taken around noon on September 1, 2023.

Of particular importance for this vignette is that three reference sufaces were
placed in the study area for the purposes of radiometric calibration. Those
surfaces were aluminum panels painted with three different colors (black, gray
and white) to allow for for differences in the temperature reached by the
panels through solar radiation. The temperature of the panels was monitored by
thermocouples plugged into CR1000X dataloggers that recorded temperature every
five seconds. This allowed surface temperature as detected by the camera to be
directly linked to high-quality temperature measurements which we will use for
radiometric calibration.

The images were downgraded from their original 640x512 resolution to 80x64 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

Image metadata can be read using the `read_metadata` function:

```{r read-metadata}
# Loading the package
library(thermal)

# Getting a vector of absolute paths to the test TIR images
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

# Reading metadata from EXIF tags
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

head(tmeta_df)
```

# Formatting data for radiometric calibration

# Rationale for radiometric calibration with `thermal`

# Fitting linear models of temperature as a function of digital number

# Quality control

# Radiometric calibration
