---
title: "Radiometric calibration with reference surfaces using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radiometric calibration with reference surfaces using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

# Introduction

In addition to correcting drift in thermal infrared (TIR) images, the `thermal`
package can also perform radiometric calibration of images or orthomosaics by
combining images with information on the temperature and position of reference
surfaces. This vignette will show a step-by-step procedure to perform such
radiometric calibration, i.e., translating digital number (DN) values returned
by a camera into actual surface temperature readings.

# Dataset

The dataset used here is based on the same flight as the vignette on thermal
drift correction, which we advise to read first. In brief, this dataset
comprises 30 TIR images taken around noon on September 1, 2023.

Of particular importance for this vignette is that three reference sufaces were
placed in the study area for the purposes of radiometric calibration. Those
surfaces are aluminum panels painted with three different colors (black, gray
and white) to allow for differences in panel temperature through absorbed solar
radiation. The temperature of the panels was monitored by type T thermocouples
plugged into CR1000X dataloggers that recorded temperature every five seconds.
This allowed thermal infrared radiation as detected by the camera to be
directly linked to high-quality temperature measurements which we will use for
radiometric calibration.

The images were downgraded from their original 640x512 resolution to 80x64 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

Image metadata can be read using the `read_metadata` function:

```{r read-metadata}
# Loading the package
library(thermal)

# Getting a vector of absolute paths to the test TIR images
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

# Reading metadata from EXIF tags
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

# Providing IDs for each picture as row names
rownames(tmeta_df) <- substr(basename(tmeta_df$SourceFile), 1, 8)

head(tmeta_df)
```

For the sake of radiometric calibration, we also provided row names to the
metadata that we will use as IDs to uniquely identify individual pictures.

# Formatting data for radiometric calibration

Radiometric calibration of TIR images requires two additional sources of data
in addition to the images themselves:

* Spatial location of the reference surfaces in the image(s)
* Time series of temperature for each of the reference surfaces

These two data sources are included in the package as datasets with names
`panels` and `temperature`, respectively. The `panels` dataset is list of `sf`
objects with indicating the coordinates of the reference surfaces (panels) in
each of the images:

```{r panels}
length(panels)
head(panels, n = 3)
```

The number of objects in the list is 27 rather than 30 (total number of images)
because three of the images did not include the panels in their frame. The
names of the list elements (DJI_0671, DJI_0677, DJI_0683, ...) link the panel
coordinates to the image that they relate to. Apart from the panel coordinates,
the `ID` column of each sf object identifies the panel that they
correspond to.

Getting the panel coordinates in the right format for radiometric calibration
(named list of sf objects, with an `ID` column for the identifier of the
surface) is probably the trickiest part of conducting radiometric calibration.
The panel coordinates of the dataset used here were obtained through a
semi-automated approach, but `thermal` also provides the function
`panel_function` to interactively derive surface coordinates from the `R`
console.

`thermal` also provides the `temperature` object as an example of surface
temperature data do be used for radiometric calibration:

```{r temperature}
str(temperature)
```

The package does not provide functions to read temperature data given the
diversity of formats in which such data may be provided. However, as long as
the following guidelines are followed for formatting temperature data,
downstream analyses should work seamlessly:

* Data is provided as a list with as many elements as the number of reference
  surfaces used.
* The name of each list element should match the `"ID"` column of the
  objects containing surface coordinates.
* Each list element should be a data.frame containing the columns `"time"`
  (vector of class POSIXct in the same time zone as the timestamps of the
  images) and `"temp"` for the temperature. Here, the temperature is provided
  in degrees Celsius, but any temperature scale should work.

The function `plot_temp` is provided to display the time series of temperature
data.


```{r plot-temp, fig.width = 8, fig.height = 6}
plot_temp(temperature,
          lcol = c("black" = "black", "gray" = "gray", "white" = "blue"),
          legend.pos = "topright")
```


Plotting the location of the panels on top of the thermal images is easy using
functionality from the `sf` and `terra` packages:

```{r plot-panels, fig.width = 8, fig.height = 6}
library(terra)
library(sf)

# We can use identifiers to index the metadata because we have set row names
terra::plot(terra::rast(tmeta_df["DJI_0671", "SourceFile"]))
plot(panels[["DJI_0671"]], col = "transparent", border = c("black", "gray", "white"), lwd = 2, add = TRUE)
```

Although image resolution is low, we can see that the black and gray panels are
warmer than the white one, as we would expect.

# Rationale for radiometric calibration with `thermal`

The `thermal` package performs radiometric calibration by assuming linear
models to compute surface temperature from thermal DN. A method for computing
such models from images and surface temperature measured by the dataloggers is
therefore necessary.

By default, the functions used to compute these linear models will summarize
the pixels corresponding to a given panel to a single value representing the
maximum density of the DN distribution. This summarized measure is less
sensitive to pixels ouside the panels than would be the mean, for example.
However, other functions can be provided to summarize (or not) the pixels
corresponding to each panel.

Using the default approach means that, in our case, linear models of
temperature will be computed from three values, one for each of the panels. The
parameters of the linear models can then be used for radiometric calibration.

# Fitting linear models of temperature as a function of digital number

From a practical standpoint, fitting linear models of surface temperature as
a function of thermal DN involves the following steps for each picture:

* Getting the temperature corresponding to each panel at the time when the
  picture was taken.
* Retrieving the values of the pixels corresponding to the panels.
* Summarizing the pixels to a single value per panel.
* Using the temperature and DN data to fit a linear model.

Many of these steps occur under the hood, but it is important to understand the
sequence of manipulations because `thermal` exposes functionality to manipulate
some of these steps through the functions `add_temp_metadata` and `thermal_lm`.

The function `add_temp_metadata` adds columns for the temperature of the
reference surfaces to the data.frame of flight metadata. Using the flight
metadata and surface temperature as input, it will create a new temperature
column for each of the reference surfaces. The temperature values are queried
such that they are the closest measurement in time to when each picture was
taken. A user-controllable threshold returns an error when no temperature
measurement is close enough (default: 10 seconds) to a picture.

```{r add-temp-metadata}
# Will return an error if any picture was taken more than 5 seconds apart from the
# closest temperature measurement
tmeta_df <- add_temp_metadata(tmeta_df, temperature, tolerance = as.difftime(5, units = "secs"))

str(tmeta_df)
```

As can be seen from the output above, three new columns have been added to the
metadata, one for the temperature of each of the three panels.

Next, the function `thermal_lm` fits the radiometric calibration models by
combining information from three sources: the thermal infrared images, the
spatial location of the reference surfaces in the images, and the temperature
of the surfaces at the time when the pictures were taken. This step requires
the row names of the flight metadata to match the IDs of the spatial polygons
representing the positions of the panels as these IDs will be used to associate
the polygons with their image.

Internally, `thermal_lm` will first extract the pixels corresponding to each
surface from the images. A summary function has to be provided to reduce the
pixels for a given panel to a single value. Typically, this function would be
`max_density`, a function that determines where the density of the distribution
is maximal. However, since the images that we use here have very low
resolution, we will use the median instead such that the analysis works even
with a single pixel. It is also possible to use different summary functions to
summarize the data for different reference surfaces (for examples, if the
distinct properties of the surfaces warrant it) by providing a named list of
functions with each name corresponding to the ID of a surface..

Once each reference surface has been reduced to a single thermal infrared value
and a single surface temperature value, `thermal_lm` will fit one linear model
of surface temperature as a function of measured thermal infrared radiation for
each picture in the dataset for which such information is available.


```{r thermal-lm}
thermal_models <- thermal_lm(metadata = tmeta_df,
			     polygons = panels,
                             summary_functions = median,
                             surfaces = c("black", "gray", "white"),
			     ncores = 2)
```

The value returned by `thermal_lm` is a list of three elements, the first of which is
the flight metadata updated with the results of the linear models for each picture.
Pictures for which no panels were in the frame display `NA` for linear model-related
variables such as slope and intercept:

```{r thermal-lm-metadata}
tmeta_df <- thermal_models$metadata
head(tmeta_df)
```

While all the necessary information for downstream analyses is embedded in that
metadata, the second element of the list contains the models themselves as well
as the pixel values that were used to fit them, for example for the first model:

```{r thermal-model-output}
thermal_models$models[[1]]
```

Finally, the third element of the returned list contains information on the
values of all the pixels that were extracted from the reference surfaces, prior
to being reduced to a single value. This can be useful for diagnosing any
issues with radiometric calibration.

# Quality control

At this stage, it may by worth assessing the results critically. We can use the
function `plot_pixtemp` to have a look at the linear models and the temperature/DN
values that they were fitted with. Black panel pixel values are shown in black,
gray panel pixel values are shown in gray, and white panel pixel values are shown
in blue. The red point show the summarized values (one per panel) that were
used to fit the models.

```{r plot-pixtemp, fig.width = 8, fig.height = 24}
par(mfrow = c(9, 3))

for(i in names(thermal_models$models)) {
        plot_pixtemp(model_data = thermal_models, id = i)
}
```

We can see that the models are a bit off for some of the images, which is to be
expected given the low resolution of the images we are using. If we were to do
any serious analysis using this data, it would be worth removing these pictures
or finding out why their results are bad.

It may also be worth looking at the shift in linear model parameters over time,
which may be an indication of thermal drift occurring. We will look at the
intercept, slope, and the predicted temperature for a relatively average DN of 3500.

```{r lm-drift, fig.width = 6, fig.height = 8}
par(mfrow = c(3, 1))

# A plot of temperature linear model intercept over time
plot(tmeta_df$DateTimeOriginal, tmeta_df$intercept, main = "Intercept", xlab = "Time", ylab = "Intercept")
lines(tmeta_df$DateTimeOriginal, tmeta_df$intercept, lty = 2)

# A plot of temperature linear model slope over time
plot(tmeta_df$DateTimeOriginal, tmeta_df$slope, main = "Slope", xlab = "Time", ylab = "Slope")
lines(tmeta_df$DateTimeOriginal, tmeta_df$slope, lty = 2)

# A plot of temperature predicted from a DN of 3500 from models fitted at various time points
tmeta_df$prediction <- thermal_predict(tmeta_df, dn_value = 3500)
plot(tmeta_df$DateTimeOriginal, tmeta_df$prediction, main = "Prediction", xlab = "Time", ylab = "Temp. predicted when DN = 3500 (°C)")
lines(tmeta_df$DateTimeOriginal, tmeta_df$prediction, lty = 2)
```

It is clear from the plot above that drift is occurring in the dataset. Indeed,
the temperature predicted from a fixed DN value varies depending on the image
that was used to fit the models, indicating instability in radiometric parameters.
If we correct the images assuming a linear model of drift, the results are
considerably improved, although not perfect:

```{r, fig.width = 6, fig.height = 6}
corrected_df <- read_metadata(thermal_files,
                              camera_tz = "Etc/GMT+5",
                              display_tz = "Etc/GMT+4",
                              tags = "minimal")

corrected_df <- correct_thermal(metadata = corrected_df,
                                correction_type = "lm",
                                output_dir = tempdir(),
                                camera_tz = "Etc/GMT+5",
                                display_tz = "Etc/GMT+4",
                                tags = "minimal",
                                nuc_threshold = 80,
                                overwrite_dst = TRUE,
                                verbose = FALSE)

rownames(corrected_df) <- substr(basename(corrected_df$SourceFile), 1, 8)

corrected_df <- add_temp_metadata(corrected_df, temperature, tolerance = as.difftime(5, units = "secs"))

corrected_df <- thermal_lm(metadata = corrected_df,
			   polygons = panels,
			   summary_functions = median,
			   surfaces = c("black", "gray", "white"),
			   ncores = 2)$metadata

# A plot of temperature predicted from a DN of 3500 from models fitted at various time points
corrected_df$prediction <- thermal_predict(corrected_df, dn_value = 3500)
plot(corrected_df$DateTimeOriginal, corrected_df$prediction, main = "Prediction", xlab = "Time", ylab = "Temp. predicted when DN = 3500 (°C)")
lines(corrected_df$DateTimeOriginal, corrected_df$prediction, lty = 2)
```

Although far from perfect, the stability of the predictions is better in the corrected data

# Radiometric calibration
