---
title: "Radiometric calibration with reference surfaces using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radiometric calibration with reference surfaces using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

# Introduction

In addition to correcting drift in thermal infrared (TIR) images, the `thermal`
package can perform radiometric calibration of images or orthomosaics by
combining images with information on the temperature and position of reference
surfaces. This vignette will show a step-by-step procedure to perform such
radiometric calibration, i.e., translating digital number (DN) values returned
by a camera into actual surface temperature readings.

# Dataset

The dataset used here is based on the same flight as the vignette on thermal
drift correction, which we advise to read first. In brief, this dataset
comprises 30 TIR images taken around noon on September 1, 2023.

Of particular importance for this vignette is that three reference sufaces were
placed in the study area during the flight for the purposes of radiometric
calibration. Those surfaces are aluminum panels painted with matte paint for
concrete of three different colors (black, gray and white) to allow for
differences in panel temperature through absorbed solar radiation. The
temperature of the panels was monitored by type T thermocouples plugged into
CR1000X dataloggers that recorded temperature every five seconds. This allowed
thermal infrared radiation as detected by the camera to be directly linked to
high-quality temperature measurements which we will use for radiometric
calibration.

The images were downgraded from their original 640x512 resolution to 80x64 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

Image metadata can be read using the `read_metadata` function:

```{r read-metadata}
# Loading the package
library(thermal)

# The functions would work without explicitly loading the sf package
# However, this will enhance the display of spatial objects in this vignette
library(sf)

# Getting a vector of absolute paths to the test TIR images
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

# Reading metadata from EXIF tags
# Setting the proper display time zone is critical here because we need to
# match the time zone of the separately recorded surface temperature data
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

# Providing IDs for each picture as row names
# This is necessary to match the panel coordinates with each picture
rownames(tmeta_df) <- substr(basename(tmeta_df$SourceFile), 1, 8)

head(tmeta_df)
```

For the sake of radiometric calibration, we also provided row names to the
metadata that we will use as IDs to uniquely identify individual pictures.

# Formatting data for radiometric calibration

Radiometric calibration of TIR images requires two additional sources of data
in addition to the images themselves:

* Spatial location of the reference surfaces in the image(s)
* Time series of temperature for each of the reference surfaces

These two data sources are included in the package as datasets with names
`panels` and `temperature`, respectively. The `panels` dataset is list of `sf`
objects with the coordinates of the reference surfaces (panels) in each of the
images:

```{r panels}
length(panels)
head(panels, n = 3)
```

The number of objects in the list is 27 rather than 30 (the total number of
images) because three of the images did not include the panels in their frame.
The names of the list elements (`DJI_0671`, `DJI_0677`, `DJI_0683`, ...) link
the panel coordinates to the image that they relate to. Apart from the panel
coordinates, the `ID` column of each sf object identifies the panel that they
correspond to.

Getting the panel coordinates in the right format for radiometric calibration
(named list of sf objects, with an `ID` column for the identifier of the
surface) is probably the trickiest part of conducting radiometric calibration.
The panel coordinates of the dataset used here were obtained through a
semi-automated approach, but `thermal` also provides the function
`create_surfaces` to interactively and graphically derive surface coordinates
from the `R` console.

`thermal` also provides the `temperature` object as an example of surface
temperature data do be used for radiometric calibration:

```{r temperature}
str(temperature)
```

The package does not provide functions to read temperature data given the
diversity of formats in which such data may be provided. However, as long as
the following guidelines are followed for formatting temperature data,
downstream analyses should work seamlessly:

* Data is provided as a list with as many elements as the number of reference
  surfaces used.
* The name of each list element should match the `"ID"` column of the
  objects containing surface coordinates.
* Each list element should be a data.frame containing the columns `"time"`
  (vector of class POSIXct in the same time zone as the timestamps of the
  images) and `"temp"` for the temperature. Here, the temperature is provided
  in degrees Celsius, but any temperature scale should work.

The function `plot_temp` is provided to display the time series of temperature
data, with one line per reference surface. By default, all lines will be black,
but here we request specific colors for each surface and add a legend to the
plot.

```{r plot-temp, fig.width = 8, fig.height = 6}
plot_temp(temperature,
          lcol = c("black" = "black", "gray" = "gray", "white" = "blue"),
          legend.pos = "topright")
```

Plotting the location of the panels on top of the thermal images is easy using
functionality from the `sf` and `terra` packages:

```{r plot-panels, fig.width = 8, fig.height = 6}
# We can use identifiers to index the metadata because we have set row names
terra::plot(terra::rast(tmeta_df["DJI_0671", "SourceFile"], noflip = TRUE))

# The argument add = TRUE plots the shapes on top of the raster
plot(panels[["DJI_0671"]],
     col = "transparent",
     border = c("black", "gray", "white"),
     lwd = 2, add = TRUE)
```

Although image resolution is low, we can see that the black and gray panels are
warmer than the white one, as we would expect.

# Rationale for radiometric calibration with `thermal`

The `thermal` package performs radiometric calibration by assuming linear
models to compute surface temperature from thermal DN. By default, the
functions used to compute these linear models will summarize the pixels
corresponding to a given panel to a single value representing the maximum
density of the DN distribution. This summarized measure is less sensitive to
pixels ouside the panels than would be the mean, for example. However, other
functions can be provided to summarize the pixels corresponding to each panel.

Using the default approach means that, in the case of the test dataset, linear
models of temperature will be computed from three values, one for each of the
panels. As we will see below, it is also possible to fit a linear model of
radiometric calibration from all images taken during a flight which include the
reference surfaces instead of from a single image.

# Basic functionality: fitting linear models of temperature as a function of digital number

From a practical standpoint, fitting linear models of surface temperature as a
function of thermal DN involves the following steps for each picture:

* Getting the temperature corresponding to each panel at the time when the
  picture was taken.
* Retrieving the values of the pixels corresponding to the panels.
* Optionally processing the distribution of DN values to summarize or filter
  them.
* Using the temperature and DN data to fit a linear model.

Many of these steps occur under the hood, but it is important to understand the
sequence of manipulations because `thermal` exposes functionality to manipulate
some of these steps through the functions `add_temp_metadata` and `thermal_lm`.

The function `add_temp_metadata` adds columns for the temperature of the
reference surfaces to the data.frame of flight metadata. Using the flight
metadata and surface temperature as input, it will create a new temperature
column for each of the reference surfaces. The temperature values are queried
such that they are the closest measurement in time to when each picture was
taken. A user-controllable threshold returns an error when no temperature
measurement is close enough (default: 10 seconds) to a picture.

```{r add-temp-metadata}
# Will return an error if any picture was taken more than 5 seconds apart from
# the closest temperature measurement
tmeta_df <- add_temp_metadata(tmeta_df, temperature, tolerance = as.difftime(5, units = "secs"))

head(tmeta_df)
```

As can be seen from the output above, three new columns have been added to the
metadata, one for the temperature of each of the three panels.

Next, the function `thermal_lm` fits the radiometric calibration models by
combining information from three sources: the thermal infrared images, the
spatial location of the reference surfaces in the images, and the temperature
of the surfaces at the time when the pictures were taken. This step requires
the row names of the flight metadata to match the IDs of the spatial polygons
representing the positions of the panels as these IDs will be used to associate
the polygons with their image. The images themselves do not need to be
explicitly loaded in memory. They will instead be read on the fly from the path
to the `SourceFile` listed in the metadata. Therefore, it is preferable to read
the metadata using absolute file paths when using `read_metadata`.

Internally, `thermal_lm` will first extract the pixels corresponding to each
surface from the images. A function has to be provided to process the
distribution of DN values obtained, for example by using the mean or median.
The values can also be filtered or left as is, in which case the radiometric
calibration model will be fitted on several values per panel. By default, the
function would be `max_density`, a function that determines where the density
of the distribution is maximal. However, since the images that we use here have
very low resolution, we will use the median instead such that the analysis
works even with a single pixel.

Once the DN values of a surface have been extracted and processed, `thermal_lm`
will fit one linear model of surface temperature as a function of measured
thermal infrared radiation for each picture in the dataset for which such
information is available. The `surfaces` argument controls which surfaces we
want to include in the computation of the models. Here, we use them all.

```{r thermal-lm}
# Parallel processing is supported through argument ncores
# However, this does not work on Windows as it uses function mclapply from the
# parallel package
thermal_models <- thermal_lm(metadata = tmeta_df,
                             polygons = panels,
                             summary_functions = median,
                             surfaces = c("black", "gray", "white"),
                             ncores = if(.Platform$OS.type == "unix") 2 else 1)
```

The value returned by `thermal_lm` is a list of three elements, the first of
which is the flight metadata updated with the results of the linear models for
each picture. Pictures for which no panels were in the frame display `NA` for
linear model-related variables such as slope and intercept:

```{r thermal-lm-metadata}
head(thermal_models$metadata)
```

While all the necessary information for downstream analyses is embedded in that
metadata, other outputs are stored to allow for quality control and
troubleshooting. For example, the second element of the list contains the
models themselves:

```{r thermal-model-output}
# Looking at the linear model for the first image
thermal_models$models[[1]]
```

Finally, the third element of the returned list contains information on the
values of all the pixels that were extracted from the reference surfaces, prior
to being processed through the summary functions. This can be useful for
diagnosing any issues with radiometric calibration.

```{r thermal-model-pixels}
head(thermal_models$pixels)
```

# Quality control and advanced functionality

At this stage, it may by worth assessing the results critically. We can use the
function `plot_pixtemp` to have a look at the linear models and the
temperature/DN values that they were fitted with. The argument `lcol` allows to
specify the colors to use for displaying the data of each reference surface
whereas the argument `id` allows to identify a vector of images to display (by
default, it will display them all).  The red point show the summarized values
(here, the median) that were used to fit the models.

```{r plot-pixtemp, fig.width = 8, fig.height = 8}
par(mfrow = c(3, 3))

# We will display data for the first 9 images
plot_pixtemp(model_data = thermal_models,
             id = names(panels)[1:9],
             lcol = c(black = "black", gray = "gray", white = "blue"))
```

We can see that the models are a bit off for some of the images, which is to be
expected given the low resolution of the images we are using. If we were to do
any serious analysis using this data, it would be worth removing these pictures
or finding out why their results are bad.

It is also possible to use different summary functions to summarize the data
for different reference surfaces (for examples, if the distinct properties of
the surfaces warrant it) by providing a named list of functions with each name
corresponding to the ID of a surface. For example, let us suppose that for some
reason we want to use the the mean for the white panel, the median for the gray
panel, and all values for the black panel. Such models can be fitted and
visualized as follows:

```{r plot-pixtemp-alternative, fig.width = 8, fig.height = 8}
alternative_models <- thermal_lm(metadata = tmeta_df,
                                 polygons = panels,
                                 summary_functions = c(white = mean, gray = median, black = identity),
                                 surfaces = c("black", "gray", "white"),
                                 ncores = if(.Platform$OS.type == "unix") 2 else 1)

par(mfrow = c(3, 3))

plot_pixtemp(model_data = alternative_models,
             id = names(panels)[1:9],
             lcol = c(black = "black", gray = "gray", white = "blue"))
```

Alternatively, we may want to fit the models using only the white and black
panels because we have found that the temperature or DN values of the gray
panel could not be trusted for this flight:

```{r plot-pixtemp-white-black, fig.width = 8, fig.height = 8}
bw_models <- thermal_lm(metadata = tmeta_df,
                        polygons = panels,
                        summary_functions = median,
                        surfaces = c("black", "white"),
                        ncores = if(.Platform$OS.type == "unix") 2 else 1)

par(mfrow = c(3, 3))

plot_pixtemp(model_data = bw_models,
             id = names(panels)[1:9],
             lcol = c(black = "black", white = "blue"))
```

Finally, we may be interested in using the data to fit a global radiometric
calibration model for the whole flight instead of using models that were
individually fitted from each image. The `global_lm` function accomplishes this
by using the output from `thermal_lm` to fit a model on the data that was
already summarized for a whole flight. The object returned by `global_lm` has
the same structure has that returned by `thermal_lm` and can thus also be
visualized with `plot_pixtemp` to see what data were used to fit the model:

```{r global-pixtemp}
global_model <- global_lm(thermal_models)

plot_pixtemp(model_data = global_model,
             lcol = c(black = "black", gray = "gray", white = "blue"))
```

# Radiometric calibration and drift

It may also be worth looking at the shift in linear model parameters over time,
which may be an indication of thermal drift occurring. We will look at the
intercept, slope, and the predicted temperature for a relatively average DN of 3500.

```{r lm-drift, fig.width = 6, fig.height = 8}
# We rename tmeta_df to the metadata stored by the output of thermal_models
tmeta_df <- thermal_models$metadata

par(mfrow = c(3, 1))

# A plot of temperature linear model intercept over time
plot(tmeta_df$DateTimeOriginal, tmeta_df$intercept, main = "Intercept", xlab = "Time", ylab = "Intercept")
lines(tmeta_df$DateTimeOriginal, tmeta_df$intercept, lty = 2)

# A plot of temperature linear model slope over time
plot(tmeta_df$DateTimeOriginal, tmeta_df$slope, main = "Slope", xlab = "Time", ylab = "Slope")
lines(tmeta_df$DateTimeOriginal, tmeta_df$slope, lty = 2)

# A plot of temperature predicted from a DN of 3500 from models fitted at various time points
tmeta_df$prediction <- thermal_predict(tmeta_df, dn_value = 3500)
plot(tmeta_df$DateTimeOriginal, tmeta_df$prediction, main = "Prediction", xlab = "Time", ylab = "Temp. predicted when DN = 3500 (°C)")
lines(tmeta_df$DateTimeOriginal, tmeta_df$prediction, lty = 2)
```

It is clear from the plot above that drift is occurring in the dataset. Indeed,
the temperature predicted from a fixed DN value varies depending on the image
that was used to fit the models, indicating instability in radiometric parameters.
If we correct the images assuming a linear model of drift, the results are
considerably improved, although not perfect:

```{r lm-correction-calibration, fig.width = 6, fig.height = 6}
corrected_df <- read_metadata(thermal_files,
                              camera_tz = "Etc/GMT+5",
                              display_tz = "Etc/GMT+4",
                              tags = "minimal")

corrected_df <- correct_thermal(metadata = corrected_df,
                                correction_type = "lm",
                                output_dir = tempdir(),
                                camera_tz = "Etc/GMT+5",
                                display_tz = "Etc/GMT+4",
                                tags = "minimal",
                                nuc_threshold = 80,
                                overwrite_dst = TRUE,
                                verbose = FALSE)

rownames(corrected_df) <- substr(basename(corrected_df$SourceFile), 1, 8)

corrected_df <- add_temp_metadata(corrected_df, temperature, tolerance = as.difftime(5, units = "secs"))

corrected_models <- thermal_lm(metadata = corrected_df,
			       polygons = panels,
			       summary_functions = median,
			       surfaces = c("black", "gray", "white"),
			       ncores = if(.Platform$OS.type == "unix") 2 else 1)

# Extracting the metadata from the output of thermal_lm
corrected_df <- corrected_models$metadata

# A plot of temperature predicted from a DN of 3500 from models fitted at various time points
corrected_df$prediction <- thermal_predict(corrected_df, dn_value = 3500)
plot(corrected_df$DateTimeOriginal, corrected_df$prediction, main = "Prediction", xlab = "Time", ylab = "Temp. predicted when DN = 3500 (°C)")
lines(corrected_df$DateTimeOriginal, corrected_df$prediction, lty = 2)
```

Although far from perfect, the stability of the predictions is better in the
corrected data

# Radiometric calibration

Finally, we can use the radiometric calibration models thus obtained to compute
surface temperature for the images acquired during the flight. We can simply use
the model coefficients of the model that we deem the best to predict the surface
temperature from the DN values of a given raster. Here, we will fit a global model
from the drift-corrected images used in the previous step:

```{r radiometric-calibration-rast, fig.width = 8, fig.height = 8}
global_corrected <- global_lm(corrected_models)

corrected_intercept <- coef(global_corrected$models$global)[1]
corrected_slope <- coef(global_corrected$models$global)[2]

par(mfrow = c(3, 3))

for(i in rownames(corrected_df)[1:9]) {
	i_rast <- terra::rast(corrected_df[i, "SourceFile"], noflip = TRUE)
	i_rast <- i_rast * corrected_slope + corrected_intercept
	terra::plot(i_rast, main = i)
}
```

