---
title: "Radiometric calibration with reference surfaces using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radiometric calibration with reference surfaces using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

# Introduction

In addition to correcting drift in thermal infrared (TIR) images, the `thermal`
package can also perform radiometric calibration of images or orthomosaics by
combining images with information on the temperature and position of reference
surfaces. This vignette will show a step-by-step procedure to perform such
radiometric calibration, i.e., translating digital number (DN) values returned
by a camera into actual surface temperature readings.

# Dataset

The dataset used here is based on the same flight as the vignette on thermal
drift correction, which we advised to read first. In brief, this dataset
comprises 30 TIR images taken around noon on September 1, 2023.

Of particular importance for this vignette is that three reference sufaces were
placed in the study area for the purposes of radiometric calibration. Those
surfaces were aluminum panels painted with three different colors (black, gray
and white) to allow for for differences in the temperature reached by the
panels through solar radiation. The temperature of the panels was monitored by
thermocouples plugged into CR1000X dataloggers that recorded temperature every
five seconds. This allowed surface temperature as detected by the camera to be
directly linked to high-quality temperature measurements which we will use for
radiometric calibration.

The images were downgraded from their original 640x512 resolution to 80x64 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

Image metadata can be read using the `read_metadata` function:

```{r read-metadata}
# Loading the package
library(thermal)

# Getting a vector of absolute paths to the test TIR images
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

# Reading metadata from EXIF tags
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

# Proviving IDs for each picture as row names
rownames(tmeta_df) <- substr(basename(tmeta_df$SourceFile), 1, 8)

head(tmeta_df)
```

For the sake of radiometric calibration, we also provided row names to the
metadata that we will used as IDs to uniquely identify individual pictures.

# Formatting data for radiometric calibration

Radiometric calibration of TIR images require two additional sources of data in
addition to the images themselves:

* Spatial location of the reference surfaces in the image(s)
* Time series of temperature for each of the reference surfaces

These two data sources are included in the package as datasets with names
`panels` and `temperature`, respectively. The `panels` dataset is list of `sf`
objects with indicating the coordinates of the reference surfaces (panels) in
each of the images:

```{r panels}
length(panels)
head(panels, n = 3)
```

The number of objects in the list is 27 rather than 30 (total number of images)
because three of the images did not include the panels in their frame. The
names of the list elements (DJI_0671, DJI_0677, DJI_0683, ...) link the panel
coordinates to the image that they relate to. Apart from the panel coordinates,
the `color` column of each sf object indicates the color of the panel that they
reference to.

Getting the panel coordinates in the right format for radiometric calibration
(named list of sf objects, with a `color` column for the identifier of the
surface) is probably the trickiest part of conducting radiometric calibration.
The panel coordinates of the dataset used here were obtained through a
semi-automated approach, but `thermal` also provides the function
`panel_function` to interactively derive surface coordinates from the `R`
console.

`thermal` also provides the `temperature` object as an example of surface
temperature data do be used for radiometric calibration:

```{r temperature}
str(temperature)
```

The package does not provide functions to read temperature data given the
diversity of formats in which such data may be provided. However, as long as
the following guidelines are followed for formatting temperature data,
downstream analyses should work seamlessly:

* Data is provided as a list with as many elements as the number of reference
  surfaces used.
* The name of each list element should match the `"color"` column of the
  objects containing surface coordinates.
* Each list element should be a data.frame containing the columns `"time"`
  (vector of class POSIXct in the same time zone as the timestamps of the
  images) and `"temp"` for the temperature. Here, the temperature is provided
  in degrees Celsius, but any temperature scale should work.

Plotting the location of the panels on top of the thermal images is easy using
functionality from the `sf` and `terra` packages:

```{r plot-panels, fig.width = 8, fig.height = 6}
library(terra)
library(sf)

# We can use identifiers to index the metadata because we have set row names
terra::plot(terra::rast(tmeta_df["DJI_0671", "SourceFile"]))
plot(panels[["DJI_0671"]], col = "transparent", border = c("black", "gray", "white"), lwd = 2, add = TRUE)
```

Although image resolution is low, we can see that the black and gray panels are
warmer than the white one, as we would expect.

# Rationale for radiometric calibration with `thermal`

# Fitting linear models of temperature as a function of digital number

# Quality control

# Radiometric calibration
