---
title: "Thermal drift correction using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thermal drift correction using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

# Introduction

The `thermal` package was designed to correct for the effect of drift in
thermal infrared (TIR) images obtained using uncooled cameras mounted on
unmanned aerial vehicles (UAVs). Indeed, the sensor arrays of these cameras are
sensitive to variation in both internal (heat produced by functioning camera)
and external (temperature, wind, radiation, and other meteorological phenomena)
conditions. As a result, data retrieved by these cameras will vary due to
factors other than the temperature of the surfaces being surveyed, which is
usually the variable that users intend to measure. This artefactual variation
in the intensity of TIR radiation recorded by cameras over time is termed
*drift*.

`thermal` provides functions for reading TIR images and their metadata,
correcting thermal drift using various underlying models, as well as a few
other functions that may be more generally useful to users of UAV-derived TIR
data. This vignette will focus on getting users started with processing TIR
data and correcting drift using various models. The underlying assumptions and
limitations of the models will be explained so that users can pick the approach
that appears to be the best for their use case.

# Test dataset

As part of this vignette, we will use a set of 30 TIR images taken around noon
on September 9, 2023. The UAV flew over a jack pine (*Pinus banksiana*) forest
but the area surveyed also includes a gravel/sand road and a parked pick-up
truck. Three aluminum panels painted with different colors (black, gray and
white) were also included to provide reference surfaces whose temperature was
measured in real time for benchmarking purposes.

The images were downgraded from their original 640x512 resolution to 80x64 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

The images can be accessed through the `system.file` function, which accesses
files shipped within a package:

```{r thermal-files}
# Loading the package
library(thermal)

# Getting a vector of absolute paths to the test TIR images
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

head(thermal_files)
```

The exact paths will differ depending on the location of the `thermal`
installation on your machine.

The first four images of the dataset look as follows:

```{r raster-images, echo = FALSE, fig.align = "center"}
par(mfrow = c(2, 2))

for(i in thermal_files[1:4]) {
	terra::plot(terra::rast(i), mar = rep(0.5, 4), legend = FALSE, axes = FALSE)
}
```

# Reading image metadata

Many operations in the `thermal` package access the images through their
metadata instead of directly through the image files. Metadata is stored within
the image files as EXIF tags which can be read through functionality provided
by the
[exifr](https://cran.r-project.org/web/packages/exifr/index.html) package. The
function `read_metadata` allows to read the metadata on a set of UAV-derived
images by providing a vector of file paths. It serves as a starting point for
any analysis involving the `thermal` package:

```{r read-metadata}
# Reading metadata from EXIF tags
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

str(tmeta_df)
```

The value returned is a data.frame with as many rows as there are images in the
dataset and various metadata variables as columns. The `tags` argument to
`read_metadata` specifies which tags should be read from the files. This can be
either a character vector of tags to read or one of a few keywords which result
in a predetermined set of tags being read. Here, the value `"minimal"`
indicates that only the tags that are deemed minimally necessary for further
processing by `thermal` should be read. By default, all tags will be read, but
this will be significantly slower as this operation is intensive in terms of
disk input/output.

The time zone arguments `camera_tz` (mandatory) and `display_tz` (optional)
influence how time will be represented in the metadata. Here, the camera wrote
the time stamps based on the Eastern Standard Time (`"Etc/GMT+5"`) but the
local time when the images were taken was actually Eastern Daylight Time
(`"Etc/GMT+4"`). Therefore, to simplify downstream analyses, we force the
metadata to be represented using this time zone.

As can be seen from the structure of the dataset above, the data.frame returned
by `read_metadata` using `tags = "minimal"` contains information about the
paths to the image files, the time when the images were taken, as well as the
orientation and geographic location of the drone when the picture was taken.
That is all the information the `thermal` package really needs to go forward,
but some users may want to extract more information depending on their
particular needs.

The `plot_metadata` function takes a dataset created with `read_metadata` and
shows a very basic layout of the locations where the flight pictures were
taken. We can see that our survey area is approximately square-shaped and
consists of eight flight lines:

```{r position-plot}
# If base_crs is not set, it will be set to WGS84 (EPSG:4326) with a warning
plot_metadata(tmeta_df, base_crs = sf::st_crs("EPSG:4326"))
```

Some parameters are available to gain more control over the output of
`plot_metadata` or plot additional features. For example, an alternative
coordinate reference system (CRS) may be specified to transform the data prior
to plotting. This can be useful to represent the coordinates in meters instead
of degrees.

```{r position-plot-crs}
plot_metadata(tmeta_df, base_crs = sf::st_crs("EPSG:4326"), new_crs = sf::st_crs("EPSG:32198"))
```

Arrows representing the trajectory of the drone can also be added:

```{r position-plot-arrows}
plot_metadata(tmeta_df, base_crs = sf::st_crs("EPSG:4326"),
              arrows = TRUE, arrow_angle = 20,
              arrow_length = grid::unit(0.15, "inches"))
```

The points can be colored according to a factor, here for example the points
are colored according to the direction that the drone is going towards:

```{r position-plot-factor}
bearing <- ifelse(abs(tmeta_df$GimbalYawDegree) > 150, "south",
                  ifelse(abs(tmeta_df$GimbalYawDegree) < 50, "north", "west"))

plot_metadata(tmeta_df, base_crs = sf::st_crs("EPSG:4326"), color = as.factor(bearing))
```

Finally, a numeric vector can also be provided as the `color` argument to map
those numeric values to a color scale:

```{r position-plot-numeric}
plot_metadata(tmeta_df, base_crs = sf::st_crs("EPSG:4326"), color = tmeta_df$GimbalYawDegree)
```

# Computing metrics on images

Once metadata has been read, several downstream analyses are possible. One that
is of particular interest in the context of thermal drift is the mean value of
all pixels in the thermal images over time. While this mean value is likely to
vary due to actual temperature differences over the area being surveyed, drift
will result in sudden or gradual changes in the mean value of the images over
time. The function `thermal_mean` accepts a data.frame of metadata and returns
a vector of numeric values corresponding to the mean value of each image.

```{r thermal-mean}
# We store the mean of each image as a new column in the metadata data.frame
tmeta_df$mean <- thermal_mean(tmeta_df)
```

While the data.frame does not contain information on the image pixels
themselves, the `SourceFile` column allows to get the path to the image. It is
therefore preferable for metadata to be read using absolute paths; this way,
the path to the images remains valid no matter what the working directory is.

By looking at the trend in the mean over time, we can identify possible drift
patterns. We can see a few sudden drops that might indicate the occurrence of
non-uniformity correction (NUC) events that resulted in the values recorded by
the camera to suddenly change. The package provides the `plot_drift` function
to view the changes in image mean over time.

```{r raw-mean-plot}
plot_drift(tmeta_df)
```

Another interesting pattern to look at is that of the mean value recorded by
each individual pixel over time. While this dataset is a bit small for that
purpose, larger datasets can provide insights into potential vignetting using
this approach. Indeed, if a large number of pictures were taken over a
relatively uniform landscape, then we would expect the average of each pixel to
be similar across the sensor array. If that is not the case, this may indicate
that some pixels consistently record higher or lower values, i.e. that
vignetting is occurring. The function `compute_vignetting` computes the
pixel-wise average over the flight and is so-named because it provides insights
into possible vignetting patterns:

```{r vignetting}
vignetting <- compute_vignetting(tmeta_df)
terra::plot(vignetting)
```

This dataset is not quite large enough to confidently identify vignetting
patterns, but the data suggest that center pixels tend to display higher values
while edge pixels display lower values.

# Drift correction

The `thermal` package provides four methods for correcting drift, each of which
has its own assumptions and limitations. In this section, we will go over each
of the correction approaches, but first we will discuss how these functions are
implemented as they share a common backbone in how the code is structured.

## Implementation of drift correction in the `thermal` package

Conceptually, drift correction aims at removing artefactual variation in
thermal values recorded by the camera while preserving variation due to actual
differences in surface temperature. Therefore, all correction approaches
implement some kind of model (statistical or empirical) that allows to
differentiate between those two components of TIR variation over time.

The function `correct_thermal` offers a unified interface for thermal drift
correction to users of the `thermal` package. While the underlying mechanics of
the correction approaches differ, all approches aim to quantify the differences
due to drift between each picture and a reference picture. These differences,
which we call *adjustment factors* in `thermal` parlance, are then added to or
subtracted from all pixels in an image provide a corrected image. That is
really all there is to it: adjustment factors are computed under a given model
and then used to adjust the values of all pixels in each image. The reference
picture can be provided explicitly by the user, but by default, the image used
as a reference will be taken from the middle of the flight (e.g. the 50th image
in a 100-image flight).

Once adjustment factors have been computed based on some model,
`correct_thermal` reads the image source files from disk, corrects them, and
saves them elsewhere on disk. The reason why `thermal` uses disk storage
instead of storing the correction results in-memory is that memory usage could
potentially exceed the RAM available on the system when processing large
datasets.  As part of the examples presented here, a temporary directory will
be used to save the results. However, in real-life usage, users should create a
directory (separate from the one where the source files are located) to store
the corrected images. `correct_thermal` will save the corrected files as TIFF
files by replacing (if necessary) the extension of the source files and saving
the corrected images to the directory provided by the user. In the current
version of `thermal`, there is no way to control output file names or format
beyond this default, but files can be renamed or converted after correction.

Importantly, EXIF metadata tags need to be transferred to the corrected image
files so that metadata (e.g. time and geographical position) are preserved.
This will be important for users aiming to perform downtream analyses such as
generating an orthomosaic from the corrected images.

Despite storing the correction results to disk instead of in-memory,
`correct_thermal` will return a data.frame of metadata on the newly corrected
data. This allows for downstream analyses to be done seamlessly on the
corrected images.

## Drift correction based on overall mean adjustement

The simplest correction model assumes that all variation in the mean of TIR
values between pictures is due to drift. While this model is almost certainly
wrong, especially if there is significant variation in land surface features
over the duration of the flight, it provides a good starting point for
discussing drift correction approaches.

Under this model, correction essentially consists in drawing a line through all
points in the plot of the mean values of thermal pictures throughout time, and
then assuming that all variation due to drift is represented by that line. This
line is indeed exactly the same as is shown using `plot_drift`:

```{r overall-model-plot}
plot_drift(tmeta_df)
```

Drift correction under the overall mean adjustment model can be run using the
following commands:

```{r overall-correction}
# Creating a temporary directory to write the corrected images to
# Normally users will use a directory on persistent storage for retrieval later
tmpdir <- tempdir()

# Running correct_thermal on the test data
# Time zone arguments need to be provided again because EXIF tags are
# transferred unmodified from the source files to the corrected files
overall_tmeta <- correct_thermal(base_data = tmeta_df,
                                 correction_type = "overall",
                                 output_dir = tmpdir,
                                 camera_tz = "Etc/GMT+5",
                                 display_tz = "Etc/GMT+4",
                                 tags = "minimal",
                                 overwrite_dst = TRUE)
```

By default, `correct_thermal` will print out messages informing users about the
adjustment factors applied to each image. Here, we can see that the image used
as a reference (the `midpoint` image), which is index 15, effectively has an
adjustment factor of 0. While we do encourage users to look at these messages
in interactive usage with real-world data, we will suppress this behaviour for
the remainder of this vignette using the argument `verbose = FALSE`.

If we look at the trend in the mean of images over time for the corrected
dataset, we can see that the mean is now nearly constant over time. This is
expected under this model because we assume that all variation in the mean
value of images is due to drift. All we did is essentially bring the mean value
of all pictures on par with that of the reference picture. By looking at the
plot, we can see minor variation in the mean of pictures over time due to the
fact that thermal values are stored on disk as integers, and some precision is
therefore lost. However, this variation (less than 1) is insignificant relative
to the scale of the measurements.

```{r overall-mean-plot}
plot_drift(overall_tmeta)
```

## Drift correction based on statistical models

The overall mean correction described above was a good introduction to the
correction framework used by `thermal`, however it is unlikely to be a good
solution for most real-world datasets as its assumptions are almost certainly
wrong. A better approach is to try and model thermal drift using statistical
models of mean thermal values over time and use those models of drift to
compute the adjustment factors.

As noted earlier, drift can occur gradually due to variation in sensor
sensitivity, but also sometimes suddenly due to NUC events triggered by the
camera firmware. Statistical models of drift need to differentiate between
those two sources of variation. To solve this problem, we chose to fit a
distinct statistical model to each flight segment occurring between two NUC
events. The rationale for this choice is that we assume that variation in
thermal values can be described statistically within the timespan of camera
settings defined by NUC, but not across different NUC settings.

This raises the question of how to define NUC events. We adopted a simple
approach that considers any difference in the mean of successive thermal images
over a fixed threshold as a NUC event. The package uses a default NUC threhsold
of 50, however, this threshold is controllable as the parameter `nuc_threshold`
in `correct_thermal` and should be assessed on a case-by-case basis depending
on the nature of the dataset.

For the purposes of this vignette, we used a threshold of 80 to define NUC
events. These are represented as red lines in the following plot, which can be
obtained by setting the `nuc_threshold` parameter in `plot_drift`:

```{r nuc-computation}
plot_drift(tmeta_df, nuc_threshold = 80)
```

Based on a threshold of 80, we can therefore see that three flight segments are
identified as separated by NUC events. Correction approaches based on the
statistical modelling of thermal drift will therefore fit three distinct models
to this data.

### Drift correction based on a linear model of thermal drift

The simplest statistical models to use for representing drift are linear
models. This approach assumes that drift varies linearly with time in-between
NUC events. The function `plot_fit` shows how a statistical model would fit the
drift pattern, here using `method = "lm"` for linear models:

```{r lm-model-plot}
plot_fit(tmeta_df, nuc_threshold = 80, method = "lm")
```

Differences between observed values and predicted values (i.e., model
residuals) are taken to represent genuine variation in TIR values and will be
preserved in the corrected data. The variation indicated by the blue line is
otherwise considered indicative of drift and will be used to compute the
adjustment factors for data correction.

Users need not to know the fine details of how to compute these linear models
and can simply provide the `correction_type = "lm"` argument to
`correct_thermal`.

```{r lm-correction}
lm_tmeta <- correct_thermal(base_data = tmeta_df,
                            correction_type = "lm",
                            output_dir = tmpdir,
                            camera_tz = "Etc/GMT+5",
                            display_tz = "Etc/GMT+4",
                            tags = "minimal",
                            nuc_threshold = 80,
                            overwrite_dst = TRUE,
                            verbose = FALSE)
```

The plot of the mean of corrected images now looks as follows:

```{r lm-mean-plot}
plot_drift(lm_tmeta)
```

This variation is somewhat more realistic than what was obtained from the
output of the overall mean model.

### Drift correction based on a cubic spline model of thermal drift

Instead of a linear model, a cubic spline can also be used to model the thermal
drift in-between NUC events. This allows for the drift model to more closely
follow variations in the data. In practice, this results in a correction that
is somewhat a hybrid between the overall mean and linear model approaches, in
that it attributes less variation to drift than the former, but more than the
latter. Setting `method = "spline"` in `plot_fit` shows how a set of splines
would fit our dataset:

```{r spline-model-plot}
plot_fit(tmeta_df, nuc_threshold = 80, method = "spline")
```

As seen above, by default, the smoothing spline will fit the data rather
closely. A more parsimonious fit can be obtained by setting the smoothing
parameter through `spline_spar`, which is eventually passed to the `spar`
parameter of function `stats::smooth.spline`. For example, `spline_spar = 0.5`
yields the following fit:

```{r spline-model-plot-spar1}
plot_fit(tmeta_df, nuc_threshold = 80, method = "spline", spline_spar = 0.5)
```

We will use this last value (`spline_spar = 0.5`) for image correction.
Applying this correction method using `correct_thermal` is fairly easy using
the `correction_type = "spline"` argument:

```{r spline-correction}
spline_tmeta <- correct_thermal(base_data = tmeta_df,
                                correction_type = "spline",
                                spline_spar = 0.5,
                                output_dir = tmpdir,
                                camera_tz = "Etc/GMT+5",
                                display_tz = "Etc/GMT+4",
                                tags = "minimal",
                                nuc_threshold = 80,
                                overwrite_dst = TRUE,
                                verbose = FALSE)
```

The plot of the mean of corrected images looks as follows after using spline
correction:

```{r spline-mean-plot}
plot_drift(spline_tmeta)
```

The variation in mean over time is fairly restricted, but not quite as much as
when using correction based on overall mean.

## Drift correction based on image overlap

The `thermal` package provides another correction method which is very
different in that it is not based on a statistical model of thermal drift.
Instead, it uses an intuitive approach to compute the adjustment factors based
on the overlap between successive pictures in a dataset.

This approach assumes that essentially no variation in surface temperature (and
therefore, in the amount of TIR radiation emitted by those surfaces) occurs in
between the few seconds that separate two successive images. Therefore, by
computing the mean value of perceived TIR in the image sections that do
overlap, it should be possible to compute just how much drift has occured
between pictures, as the surface captured is the same in both sections.

### Finding the overlaps between images

The correction based on image overlap is by far the most computing-intensive
method as it first needs to determine the sections of images that overlap. To
find these overlapping sections, the `thermal` package needs a formal
description of how the coordinates of an image can be translated to the
coordinates of a neighbouring one using a geometric transformation that
combines a rotation and a translation. This is modeled using three
transformation parameters:

* *theta*, the angle of rotation (in degrees) between sucessive pictures
* *htrans*, the horizontal translation (in pixels) between successive pictures
* *vtrans*, the vertical translation (in pixels) between successive pictures

These transformation parameters need to be estimated for each pair of
successive pictures in a dataset. Therefore, for the 30 pictures in our
dataset, 29 sets of transformation parameters need to be estimated.

The function `optimize_transform` provides the functionality to compute those
parameters. Under the hood, it first tries various parameter combinations based
on a brute-force approach and then passes on the best parameter combination to
`stats::optim` to refine the estimates. The metric that it aims to optimize is
the correlation between the pixel values of two pictures for which we want to
find the transformation parameters. Many parameters are available to control
the optimization procedure and we highly encourage users who intend to use this
approach to carefully read the function's documentation. Here, we will simply
show a glimpse of the available functionality. The next code chunk shows the
estimation of the transformation parameters for the first two pictures in the
dataset.

```{r, optimize-example}
# We can provide a first guess of what the rotation will be based on the
# difference in gimbal yaw angle recorded by the camera at the two time points
theta_guess <- -(tmeta_df[2, "GimbalYawDegree"] - tmeta_df[1, "GimbalYawDegree"])

trans_params <- optimize_transform(x = terra::rast(tmeta_df[1, "SourceFile"]),
                                   y = terra::rast(tmeta_df[2, "SourceFile"]),
                                   theta1 = theta_guess,
                                   reltol = 10^-3)
```

The package provides the `check_transform` function to visually verify the
accuracy of transformation parameters. This function works best in interactive
mode, where the user is queried for clicks on the source image and the location
of the transformed coordinates is drawn on the target image. This way, the
location of features of interest can be verified. Otherwise, a number of
coordinates are randomly sampled and shown using distinct colors on both
images. In all cases, the oultine of the images relative to each other are
shown using a red box. In the following plot, we can see that the
transformation parameters are reasonably accurate.

```{r check-overlaps, fig.width = 8, fig.height = 6}
# Setting the seed for reproducibility of the vignette
set.seed(1)

# Verifying the transformation parameters non-interactively
check_transform(x = terra::rast(tmeta_df[1, "SourceFile"]),
                y = terra::rast(tmeta_df[2, "SourceFile"]),
                params = trans_params, n = 40)
```

We will need the transformation parameters for the whole dataset to correct
based on image overlap. Here, we aggressively use some parameters offered by
`optimize_transform` to speed up code execution for the purpose of generating
the vignette. The parameter `min_cor = 0.6` allows to pass control to
`stats::optim` earlier in the search for the best parameters, while `fact = 4`
aggregates the raster by a factor of four in the brute-force search step. The
parameter `reltol = 10^-2` also stops the `stats::optim` routine earlier in its
search for optimal parameters. All three parameters speed up the search
greatly, at the possible cost of not finding the right transformation
parameters. Under normal circumstances, we advise to use more stringent
parameters. Furthermore, we allow two cores to perform some of the
computations in parallel.  However, this option is not available on Windows
systems as it relies on `parallel::mclapply`.

Instead of using `optimize_transform`, here we use the convenient wrapper
`compute_overlaps` which computes transformation parameters for all images in
a given flight and adds the resulting information as metadata columns that
are needed for overlap correction. Here, we can indeed see that the columns
`theta`, `htrans`, `vtrans` and `corr` have been added to the metadata. The
first value of each of these columns is `NA` because there are no transformation
parameters linking the first image in the dataset to a previous image.

```{r optimize-all}
# Adding the transformation parameters to the metadata
tmeta_df <- compute_overlaps(tmeta_df,
                             min_cor = 0.6,
                             fact = 4,
                             cores = if(.Platform$OS.type == "unix") 2 else 1,
                             verbose = FALSE,
                             reltol = 10^-2)

str(tmeta_df)
```

The `corr` column is not necessary for image correction but it provides a
useful diagnostic for the quality of the overlap parameters. It corresponds to
the correlation between the values of two successive images when applying the
optimized transformation. The histrogram below shows the distribution of
correlation values of all coordinate transformations for our test flight. The
quality of the transformation parameters is rather good as most values are > 0.8.

```{r corr-hist, fig.width = 6, fig.height = 6}
hist(tmeta_df$corr, breaks = 10, main = "Histogram of image overlap correlation values",
     ylab = "Count", xlab = "Correlation")
```

Low values (for example, below 0.75) should raise suspicion and warrant a
detailed look at the coordinate transformation, for example using
`check_transform`. Here, a closer look at our image pair with the lowest
correlation (`r round(min(tmeta_df$corr, na.rm = TRUE), 2)`) suggests that the
transformation parameters are indeed a bit off for that image. This indicates
that more stringent parameters should be used when computing the transformation
parameters. Here, it is not surprising that we struggle in finding good
parameters since we are working with low-resolution images. Real-world datasets
will not commonly suffer from this problem.

```{r check-overlaps-mincorr, fig.width = 8, fig.height = 6}
# Setting the seed for reproducibility of the vignette
set.seed(1)

# Finding the image index with the lowest correlation
min_image <- which.min(tmeta_df$corr)

# Verifying the transformation parameters non-interactively
check_transform(x = terra::rast(tmeta_df[min_image - 1, "SourceFile"]),
                y = terra::rast(tmeta_df[min_image, "SourceFile"]),
                params = trans_params, n = 40)
```

### Correction based on image overlap

Once the transformation parameters are known and have been added to the
metadata, `correct_thermal` can be invoked with `correction_type = "overlap"`.
Instead of relying on a statistical model, this approach will use the
transformation parameters computed above to determine overlapping sections of
successive images. Adjustment factors will then be computed based on the
difference between the mean of the images in the region where they overlap.

```{r overlap-correction}
overlap_tmeta <- correct_thermal(base_data = tmeta_df,
                                 correction_type = "overlap",
                                 output_dir = tmpdir,
                                 camera_tz = "Etc/GMT+5",
                                 display_tz = "Etc/GMT+4",
                                 tags = "minimal",
                                 overwrite_dst = TRUE,
                                 verbose = FALSE)
```

The plot of the mean of images in the corrected looks as follows after using
overlap correction:

```{r overlap-mean-plot}
plot_drift(overlap_tmeta)
```

As a cautionary note, we have found the overlap method to work well along
straight flight lines, but not so much when the drone makes frequent turns. We
suspect that this is because when the drone turns, the overlapping regions in
successive images then come from very different locations on the sensor array.
If the camera suffers from vignetting issues, the comparison of the overlapping
sections of the images will be biased and error propagation becomes likely. In
this case, overlap correction may worsen the results compared to no correction
at all.

# Session info

This vignette was produced under the following system:

```{r session-info}
sessionInfo()
```

