---
title: "Thermal drift correction using the thermal package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thermal drift correction using the thermal package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center")
```

```{r packages}
library(thermal)
library(terra)
```

# Introduction

The `thermal` package was designed to correct for the effect of drift in
thermal infrared (TIR) images obtained using uncooled cameras mounted on
unmanned aerial vehicles (UAVs). Indeed, the sensor arrays of these cameras are
sensitive to variation in both internal (heat produced by functioning camera)
and external (temperature, wind, radiation, and other meteorological phenomena)
conditions. As a result, data retrieved by these cameras will vary due to
factors other than the temperature of the surfaces being surveyed, which is the
usually the metric that users intend to measure. This artefactual variation in
the intensity of TIR radiation recorded by cameras ofter time is termed
*drift*.

`thermal` provides functions for reading TIR images as well as their metadata,
correct thermal drift using various underlying models, as well as a few other
functions that may be more generally useful to users of UAV-derived TIR data.
This vignette will focus on getting users started with processing TIR data and
correcting drift using various models. The underlying assumptions and
limitations of the models will be explained to that users can pick the approach
that appears to be the best for their use case.

# Test dataset

As part of this vignette, we will use a set of 90 TIR images taken during a
around noon on September 9, 2023. The UAV flew over a jack pine (*Pinus
banksiana*) forest but the area surveyed also includes a gravel/sand road and a
parked pick-up truck. Three aluminum panels painted with different colors
(black, gray and white) were also included to provide reference surfaces whose
surface temperature was measured in real time for benchmarking purposes.

The images were downgraded from their original 640x512 resolution to 160x128 to
reduce their size for shipping along with the package and speed up processing
for testing purposes. While the quality of the images is reduced, this is of
little relevance to the examples presented in this vignette.

The images can be accessed through the `system.file` function, which accesses
files shipped within a package:

```{r thermal-files}
thermal_files <- dir(system.file("extdata", package = "thermal"),
                     pattern = "thermal\\.tiff$", full.names = TRUE)

head(thermal_files)
```

The exact paths will differ depending on the location of the `thermal`
installation on your machine.

The first four images of the dataset look as follows:

```{r raster-images, echo = FALSE, fig.align = "center"}
par(mfrow = c(2, 2))

for(i in thermal_files[1:4]) {
	plot(rast(i), mar = rep(0.5, 4), legend = FALSE, axes = FALSE)
}
```


# Reading image metadata

Many operations in the `thermal` package access the images through their
metadata instead of directly through the image files. Metadata is stored within
the image files in as EXIF tags which can be read through functionality
provided by the
[exifr](https://cran.r-project.org/web/packages/exifr/index.html) package. The
function `read_metadata` allows to read the metadata on a set of UAV-derived
images by providing a vector of file paths and serves as a starting point for
any analysis involving the `thermal` package:

```{r read-metadata}
# Reading metadata from EXIF tags
tmeta_df <- read_metadata(thermal_files,
                          camera_tz = "Etc/GMT+5",
                          display_tz = "Etc/GMT+4",
                          tags = "minimal")

str(tmeta_df)
```

The value returned is a data.frame with as many rows as there are images in the
dataset and various metadata variables as columns. The `tags` argument to
`read_metadata` specifies which tags should be read from the files. This can be
either a character vector of tags to read or one of a few keywords which result
in a predetermined set of tags being read. Here, the value `"minimal"`
indicates that only the tags that are deemed minimally necessary for further
processing by `thermal` should be read. By default, all tags will be read, but
this will be significantly slower as this operation is intensive in terms of
disk input/output.

The time zone arguments `camera_tz` (mandatory) and `display_tz` (optional)
influence how time will be represented in the metadata. Here, the camera wrote
the time stamps based on the Eastern Standard Time (`"Etc/GMT+5"`) but the
local time when the images were taken was actually Eastern Daylight Time
(`"Etc/GMT+4"`). Therefore, to simplify downstream analyses, we force the
metadata to be represented using this time zone.

As can be seen from the structure of the dataset above, the data.frame returned
by `read_metadata` using `tags = "minimal"` contains information about the
paths to the image files, the time when the images were taken, as well as the
orientation and geographical localization of the drone when the picture was taken.
That is all the information the `thermal` package really needs to go forward,
but some users may want to extract more information depending on their particular
needs.

The `position_plot` function takes a dataset read by `read_metadata` as input
and shows a very basic layout of the locations where the flight pictures were
taken. We can see that our survey area is approximately square-shaped and
consists of eight flight lines:

```{r position-plot}
position_plot(tmeta_df)
```

# Computing metrics on images

Once metadata has been read, several downstream analyses are possible. One that
is of particular interest in the context of thermal drift is the mean value of
all pixels in the thermal images over time. While this mean value is likely to
vary due actual temperature variation of the surface being sueveyed, drift will
result in sudden or gradual changes in the mean value of the images over time.
The function `thermal_mean` accepts a data.frame of metadata and returns a
vector of numeric values corresponding to the mean value of each image.

```{r thermal-mean}
# We store the mean of each image as a new column in the metadata data.frame
tmeta_df$mean <- thermal_mean(tmeta_df)
```
 While the data.frame does not contain information on the image pixels
themselves, the `SourceFile` column allows to get the path to the image. It is
therefore preferable for metadata to be read using absolute paths; this way, the
path to the images remains valid no matter what the working directory is.

By looking at the trend in the mean over time, we can identify possible drift
patterns. We can see a few sudden drops that might indicate the occurrence of
non-uniformity correction (NUC) events that resulted in the values recorded be
the camera to suddenly change.

```{r raw-mean-plot}
plot(tmeta_df$DateTimeOriginal, tmeta_df$mean)
lines(tmeta_df$DateTimeOriginal, tmeta_df$mean)
```

Another interesting pattern to look at is that of the mean value recorded by
each individual pixel over time. While this dataset is a bit small for that
purpose, larger datasets can provide insights into potential vignetting using
this approach. Indeed, if a large number of pictures were taken of a relatively
uniform landscape, then we would expect the average of each pixel to be similar
across the sensor array. If that is not the case, this may indicate that some
pixels consistently record higher or lower values, i.e. that vignetting is
occurring. The function `compute_vignetting` computes the pixel-wise average
over the flight and is so-named because it provides insights into possible
vignetting patterns:

```{r vignetting}
vignetting <- compute_vignetting(tmeta_df)
plot(vignetting)
```

Here, the dataset is not quite large enough to confidently identify vignetting
patterns, but the data suggest that center pixels tend to be have higher values
while edge pixels have lower values.

# Drift correction

The `thermal` package provides four methods for correcting drift, each of which
has its own assumptions and limitations. In this section, we will go over each
of the correction approaches, but first we will discuss how these functions are
implemented as they share a common backbone in how the code is structured.

## Implementation of drift correction in the `thermal` package

Conceptually, drift correction aims at removing artefactual variation in
thermal values recorded by the camera while preserving variation due to actual
differences in surface temperature. Therefore, all correction approaches
implement some kind of model (statistical or empirical) that allows to
differentiate between those two components of TIR variation over time.

The function `correct_thermal` offers a unified interface for thermal drift
correction to users of the `thermal` package. While the underlying mechanics of
the correction approaches differ, all approches aim to quantify the differences
due to drift between each picture and a reference picture. These differences,
which we call *adjustment factors* in `thermal` parlance, are then added
or subtracted from all pixels in an image provide a corrected image. That is
really all there is to it: adjustment factors are computed under a given model
and then used to adjust the values of all pixels in each image. The reference
picture that the adjustment factors are relative to can be provided explicitly
by the user; by default, the image used as a reference will be taken from
the middle of the flight (e.g. the 50th image in a 100-image flight).

Once adjustment factors have been computed based on some model,
`correct_thermal` reads the image source files from disk, corrects them, and
saves them elsewhere on disk. The reason why `thermal` uses disk storage
instead of storing the correction results in-memory is that memory usage could
potentially exceed the RAM available on the system when processing large
datasets.  As part of the examples presented here, a temporary directory will
be used to save the results. However, in real-life usage, users should create a
directory (separate from that where the source files are located) to store the
corrected images. `correct_thermal` will save the corrected files as TIFF files
by replacing (if necessary) the extension of the source files and saving the
corrected images to the directory provided as input by the user. In the current
version of `thermal`, there is no way to control output file names or format
beyond this default.

Importantly, EXIF metadata tags need to be transferred to the corrected image
files so that metadata (e.g. time and geographical position) are preserved.
This will be important for users aiming to perform downtream analyses such as
generating an orthomosaic of the corrected images.

Despite storing the correction results to disk instead of in-memory,
`correct_thermal` will return a data.frame of metadata on the newly corrected
data. This allows for downstream analyses to be done seamlessly on the
corrected images.

## Drift correction based on overall mean adjustement

The simplest correction model assumes that all variation in the mean of TIR
values between pictures is due to drift. While this model is almost certainly
wrong, especially if there is significant variation in land surface features
over the duration of the flight, it provides a good starting point for
discussing drift correction approaches.

Under this model, correction essentially consists in drawing a line through all
points in the plot of the mean values of thermal pictures throughout time, and
then assuming that all variation due to drift is represented by that line:

```{r overall-model-plot}
plot(tmeta_df$DateTimeOriginal, tmeta_df$mean)
lines(tmeta_df$DateTimeOriginal, tmeta_df$mean, col = "blue", lty = 2)
```

Drift correction under the overall mean adjustment model can be run using the
following commands:

```{r overall-correction}
# Creating a temporary directory to write the corrected images to
# Normally users will use a directory on persistent storage for retrieval later
tmpdir <- tempdir()

# Running correct_thermal on the test data
# Time zone arguments need to be provided again because EXIF tags are
# transferred unmodified from the source files to the corrected files
overall_tmeta <- correct_thermal(base_data = tmeta_df,
                                 correction_type = "overall",
                                 output_dir = tmpdir,
                                 camera_tz = "Etc/GMT+5",
                                 display_tz = "Etc/GMT+4",
                                 tags = "minimal",
                                 overwrite_dst = TRUE)
```

By default, `correct_thermal` will print out messages informing users about the
adjustment factors applied to each image. Here, we can see that the image used
as a reference (the `midpoint` image), which is index 45, effectively has an
adjustment factor of 0. While we do encourage users to look at these messages
in interactive usage with real-world data, we will suppress this behaviour for
the remainder of this report using the argument `verbose = FALSE`.

If we look at the trend in the mean of images over time for the corrected
dataset, we can see that the mean is now nearly constant over time. This is
expected under this model because we assume that all variation in the mean
value of images is due to drift. All we did is essentially bring the mean value
of all pictures on par with that of the reference picture. By looking at the
plot, we can see minor variation in the mean of pictures over time due to the
fact that thermal values are stored on disk as integers, and some precision is
therefore lost.  However, this variation (less than 1) is insignificant
relative to the scale of the measurements.

```{r overall-mean-plot}
overall_tmeta$mean <- thermal_mean(overall_tmeta)
plot(overall_tmeta$DateTimeOriginal, overall_tmeta$mean)
lines(overall_tmeta$DateTimeOriginal, overall_tmeta$mean)
```

## Drift correction based on statistical models

The overall mean correction described above was a good introduction to the
correction framework used by `thermal`, however it is unlikely to be a good
solution for most real-world datasets as its assumptions are almost certainly
wrong. A better approach is to try and model thermal drift using statistical
models of mean thermal values over time and use those models of drift to
compute the adjustment factors.

As noted earlier in this report, drift can occur gradually due to variation in
sensor sensitivity, but also sometimes suddenly due to NUC events triggered by
the camera firmware. Statistical models of drift need to differentiate between
those two sources of variation. To solve this problem, in the `thermal` package,
we chose to fit a distinct statistical model to each flight segment that occurred
between two NUC events. The rationale for this choice is that we assume that
variation in thermal values can be described statistically within the timespan
of camera settings defined by NUC, but not across different NUC settings.

This raises the question of how we can define NUC events. We adopted a simple
approach that considers any variation in the mean of thermal images over 50
as a NUC event, as this constitutes a fairly large change from one picture to
the next. However, this threshold is controllable as the parameter `nuc_threshold`
in `correct_thermal` and should be assessed on a case-by-case basis depending on
the nature of the dataset.

For our test dataset, the NUC events based on a threshold of 50 are represented
as red lines in the following plot:

```{r nuc-computation}
# Declaring NUC events when the difference between successive pictures is > 50
# The last picture cannot be linked to NUC because there is no picture after it
tmeta_df$nuc <- c(abs(diff(tmeta_df$mean)) > 50, FALSE)

# Plotting the data, with NUC events as red lines
plot(tmeta_df$DateTimeOriginal, tmeta_df$mean)
for(i in which(tmeta_df$nuc)) lines(tmeta_df[c(i, i + 1), "DateTimeOriginal"], tmeta_df[c(i, i + 1), "mean"], col = "red")
```

Based on a threshold of 50, we can therefore see that three flight segments are
identified as separated by NUC events. Correction approaches based on the
statistical modelling of thermal drift will therefore fit three different
models to this data.

### Drift correction based on a linear model of thermal drift

The simplest statistical models to use for representing drift are linear
models. This approach assumes that drift varies linearly with time in-between
NUC events. The code and plot below shows how `correct_thermal` works under the
hood to model thermal drift based on these assumptions:

```{r lm-model-plot}
# We re-compute NUC events a bit differently so we can group data points
# according to what NUC settings they belong to
tmeta_df$nuc <- c(FALSE, abs(diff(tmeta_df$mean)) > 50)
tmeta_df$nuc_group <- cumsum(tmeta_df$nuc)

# We need to adjust the values of DateTimeOriginal to ensure that a fit is found
tmeta_df$time <- tmeta_df$DateTimeOriginal - min(tmeta_df$DateTimeOriginal)

# We compute the predicted values based on linear models
lm_predicted <- lapply(split(tmeta_df, tmeta_df$nuc_group), function(x) {
			       fitted(lm(mean ~ time, data = x))
				 })

lm_predicted <- unlist(lm_predicted)

plot(tmeta_df$DateTimeOriginal, tmeta_df$mean)
lines(tmeta_df$DateTimeOriginal, lm_predicted, col = "blue", lty = 2)
```

Differences between observed values and predicted values (i.e., model
residuals) are taken to represent genuine variation in TIR values and will be
preserved in the corrected data. The variation indicated by the blue line is
otherwise considered indicative of drift and will be used to compute the
adjustment factors for data correction.

Users need not to know the fine details of how to compute these linear
models and can simply provide the `correction_type = "lm"` argument to
`correct_thermal`.

```{r lm-correction}
lm_tmeta <- correct_thermal(base_data = tmeta_df,
                            correction_type = "lm",
                            output_dir = tmpdir,
                            camera_tz = "Etc/GMT+5",
                            display_tz = "Etc/GMT+4",
                            tags = "minimal",
                            nuc_threshold = 50,
                            overwrite_dst = TRUE,
                            verbose = FALSE)
```

The plot of the mean of corrected images now looks as follows:

```{r lm-mean-plot}
lm_tmeta$mean <- thermal_mean(lm_tmeta)
plot(lm_tmeta$DateTimeOriginal, lm_tmeta$mean)
lines(lm_tmeta$DateTimeOriginal, lm_tmeta$mean)
```

This variation is somewhat more realistic than what was obtained from the
output of the overall mean model.

### Drift correction based on a cubic spline model of thermal drift

Instead of a linear model, a cubic spline can also be used to model the thermal
drift in-between NUC events. This allows for the drift model to more closely
follow variations in the data. In practice, this results in a correction that
is somewhat a hybrid between the overall mean and linear model approaches, in
that it attributes less variation to drift than the former, but more than the
latter.

The code and plot below shows how `correct_thermal` works under the hood
to model thermal drift based on cubic splines:

```{r spline-model-plot}
# We compute the predicted values based on cubic splines
spline_predicted <- lapply(split(tmeta_df, tmeta_df$nuc_group), function(x) {
			    fitted(smooth.spline(x$time, x$mean))
		    })

spline_predicted <- unlist(spline_predicted)

plot(tmeta_df$DateTimeOriginal, tmeta_df$mean)
lines(tmeta_df$DateTimeOriginal, spline_predicted, col = "blue", lty = 2)
```

Again, applying this correction method using `correct_thermal` is fairly
easy using the `correction_type = "spline"` argument:

```{r spline-correction}
spline_tmeta <- correct_thermal(base_data = tmeta_df,
                                correction_type = "spline",
                                output_dir = tmpdir,
                                camera_tz = "Etc/GMT+5",
                                display_tz = "Etc/GMT+4",
                                tags = "minimal",
                                nuc_threshold = 50,
                                overwrite_dst = TRUE,
                                verbose = FALSE)
```

The plot of the mean of corrected images looks as follows after using spline
correction:

```{r spline-mean-plot}
spline_tmeta$mean <- thermal_mean(spline_tmeta)
plot(spline_tmeta$DateTimeOriginal, spline_tmeta$mean)
lines(spline_tmeta$DateTimeOriginal, spline_tmeta$mean)
```

The variation in mean over time is fairly restricted, but not quite as much as
when using correction based on overall mean.

## Drift correction based on image overlap

The `thermal` package provides another correction method which is very
different in that it is not based on a statistical model of thermal drift.
Instead, it uses an intuitive approach to compute the adjustment factors based
on the overlap between successive pictures in a dataset.

This approach assumes that essentially no variation in surface temperature (and
therefore, in the amount of TIR emitted by those surfaces) occur in between the
few seconds that separate two pictures taken one after the other. Therefore, by
computing the mean value of perceived TIR in the image sections that do
overlap, it should be possible to compute just how much drift has occured
between pictures, as the surface captured is the same in both pictures.

### Finding the overlaps between images

The correction based on image overlap is by far the most computing-intensive
method as it first needs to determine the sections of images that overlap. To
find these overlapping sections, the `thermal` package needs a formal
description of how the coordinates of an image can be translated to the
coordinates of a neighbouring one in space using a geometric transformation
that combines a rotation and a translation. This is modeled using three transformation
parameters:

* *theta*, the angle of rotation (in degrees) between sucessive pictures
* *htrans*, the horizontal translation (in pixels) between successive pictures
* *vtrans*, the vertical translation (in pixels) between successive pictures

These transformation parameters need to be estimated for each pair of
successive pictures in a dataset. Therefore, for the 90 pictures in our
dataset, 89 sets of transformation parameters need to be estimated.

The function `optimize_transform` provides the interface to compute those
parameters. Under the hood, it first tries various parameter combinations based
on a brute-force approach and then passes on the best parameter combination to
`stats::optim` to refine the estimates. The metric that it aims to optimize is
the correlation between the pixel values of two pictures for which we want to
find the transformation parameters. Many parameters are available to control
the optimization procedure and we highly encourage users who intend to use this
approach to carefully read the function's documentation. Here, we will simply
show a glimpse of the available functionality. The next code chunk shows the
estimation of the transformation parameters for the first two pictures in the
dataset.

```{r, optimize-example}
# We can provide a first guess of what the rotation will be based on the
# difference in gimbal yaw angle recorded by the camera at the two time points
theta_guess <- -(tmeta_df[2, "GimbalYawDegree"] - tmeta_df[1, "GimbalYawDegree"])

trans_params <- optimize_transform(x = terra::rast(tmeta_df[1, "SourceFile"]),
                                   y = terra::rast(tmeta_df[2, "SourceFile"]),
                                   theta1 = theta_guess,
                                   reltol = 10^-3)
```

Here we will show what the overlaps look like.

We will need the transformation parameters for the whole dataset to correct
based on image overlap. Here, we aggressively use some parameters offered by
`optimize_transform` to speed up code execution for the purpose of generating
the vignette. The parameter `min_cor = 0.6` allows to pass control to
`stats::optim` earlier in the search for the best parameters, while `fact = 4`
aggregates the raster by a factor of four in the brute-search step. The
parameter `reltol = 10^-2` also stops the `stats::optim` routine earlier in its
search for optimal parameters. All three parameters speed up the search
greatly, at the possible cost of not finding the right transformation
parameters. Under normal production conditions, we would advise to use more
stringent parameters.  Furthermore, we allow the use of four cores to perform
some of the computations in parallel.  However, this option is not available on
Windows systems as it relies on `parallel::mclapply`.

```{r optimize-all}
# This list will store the output of optimize transform for each image pair
tparams <- list()

for(i in 2:nrow(tmeta_df)) {
	# Computing the initial guess for theta
	theta_guess <- -(tmeta_df[i, "GimbalYawDegree"] - tmeta_df[i - 1, "GimbalYawDegree"])

	# Filling the tparams list with the parameter estimates
	tparams[[i - 1]] <- optimize_transform(x = terra::rast(tmeta_df[i - 1, "SourceFile"]),
                                               y = terra::rast(tmeta_df[i, "SourceFile"]),
                                               theta1 = theta_guess,
                                               min_cor = 0.6,
                                               fact = 4,
                                               cores = if(.Platform$OS.type == "unix") 4 else 1,
                                               verbose = FALSE,
                                               reltol = 10^-2)

}
```

We can have a look at the distribution of correlation values and check the
overlap for the image pair with the lowest correlation.

### Correction based on image overlap

Once the transformation parameters are known and have been added to the
metadata, `correct_thermal` can be invoked with `correction_type = "overlap"`.
Instead of relying on a statistical model, this approach will use the
transformation parameters computed above to determine the section of each image
pair that overlap with each other. Adjustment factors will then be computed
based on the difference between the mean of the images in the region where they
overlap. In addition to the usual arguments, we also need to provide the
`tparams` parameter, which contains the list of transformation parameters
computed above.

```{r overlap-correction}
overlap_tmeta <- correct_thermal(base_data = tmeta_df,
                                 correction_type = "overlap",
                                 output_dir = tmpdir,
                                 camera_tz = "Etc/GMT+5",
                                 display_tz = "Etc/GMT+4",
                                 tags = "minimal",
                                 overwrite_dst = TRUE,
                                 tparams = tparams,
				 verbose = FALSE)
```

The plot of the mean of images in the corrected looks as follows after using
overlap correction:

```{r overlap-mean-plot}
overlap_tmeta$mean <- thermal_mean(overlap_tmeta)
plot(overlap_tmeta$DateTimeOriginal, overlap_tmeta$mean)
lines(overlap_tmeta$DateTimeOriginal, overlap_tmeta$mean)
```

As a cautionary note, we have found the overlap method to work well along
straight flight lines, but not so much when the drone makes frequent turns. We
suspect that this is because when the drone turns, the overlapping regions in
successive images then come from very different locations on the sensor array.
If the camera suffers from vignetting issues, the comparison of the overlapping
sections of the images will be biased and error propagation becomes likely. In
this case, overlap correction may worsen the results compared to no correction
at all.

# Session info

This vignette was produced under the following system:

```{r session-info}
sessionInfo()
```

